How to Avoid OOTA Without Really Trying

### 2024-09-12 paper/ootavacuous.tex
ootavacuous: Consistently use "per-thread analysis"
Reported-by: Alan Stern <stern@rowland.harvard.edu>
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

The definition given below is applicable only to C++ implementations
that treat all atomic objects as though they are volatile or
quasi volatile.
(For compilers that perform only single-thread analysis, not global
analysis, quasi volatile is sufficient.)
In this setting we can relate abstract and hardware executions by
means of the map of accesses described in
Section~\ref{sec:Volatile and Quasi Volatile Accesses}.
The key insight is that this allows us to consider semantic
dependencies at the level of the machine code, where they are much
more tractable.
This insight allows us to demonstrate that OOTA cycles cannot form
in the C++ implementations addressed by this paper.

@@

In this section we consider implementations whose compilers perform
only single-thread analysis and treat atomic objects as
quasi volatile.
This implies that if two different programs contain the same thread
(i.e., the same source code for the functions and objects in the thread), the
machine code generated by the compiler for the thread will be the
same in the two programs.

@@

A corollary of this result is that if an implementation's compiler
either
\begin{itemize}
\item	uses single-thread analysis and treats atomic objects as
	quasi volatile, or
\item	uses global analysis and treats atomic objects as volatile,
\end{itemize}
then programs produced by that implementation will never exhibit OOTA.
Thus the implementation will automatically comply with full C++, even
though it may be been designed only to comply with loose C++.

@@

However, a single-thread analysis compiler can translate this into the
machine-code equivalent of:
\begin{quote}
\begin{verbatim}
int r1 = (x != 0);
int r2;
int r3 = (x != 0);
int r4 = (y != 0);
if (r1 != r3) {
   r2 = r4;
   z = 1;
} else {
   r2 = r1;
   z = (r2 == r4);
}
\end{verbatim}
\end{quote}
In effect, the \co{r2} load is merged with the \co{r1} load in
executions where \co{r1} is equal to \co{r3} or to \co{r4}, and it is
merged with the \co{r3} load in other executions.

@@

This paper focused solely on compilers, primarily those that do only
local per-thread anaylsis and optimization.
It introduced the notion of quasi-volatile object accesses and gave
a definition for semantic dependency that is relative to a particular
C++ implementation.
It formulated a Fundamental Property of semantic dependency and proved
that implementations subject to some very mild restrictions on how
they treat atomic object accesses do have this property.
Finally, it demonstrated that satisfying the Fundamental Property
guarantees an implementation cannot give rise to OOTA executions.

>>>>

The definition given below is applicable only to C++ implementations
that treat all atomic objects as though they are volatile or
quasi volatile.
(For compilers that perform only per-thread analysis, not global
analysis, quasi volatile is sufficient.)
In this setting we can relate abstract and hardware executions by
means of the map of accesses described in
Section~\ref{sec:Volatile and Quasi Volatile Accesses}.
The key insight is that this allows us to consider semantic
dependencies at the level of the machine code, where they are much
more tractable.
This insight allows us to demonstrate that OOTA cycles cannot form
in the C++ implementations addressed by this paper.

@@

In this section we consider implementations whose compilers perform
only per-thread analysis and treat atomic objects as
quasi volatile.
This implies that if two different programs contain the same thread
(i.e., the same source code for the functions and objects in the thread), the
machine code generated by the compiler for the thread will be the
same in the two programs.

@@

A corollary of this result is that if an implementation's compiler
either
\begin{itemize}
\item	uses per-thread analysis and treats atomic objects as
	quasi volatile, or
\item	uses global analysis and treats atomic objects as volatile,
\end{itemize}
then programs produced by that implementation will never exhibit OOTA.
Thus the implementation will automatically comply with full C++, even
though it may be been designed only to comply with loose C++.

@@

However, a per-thread analysis compiler can translate this into the
machine-code equivalent of:
\begin{quote}
\begin{verbatim}
int r1 = (x != 0);
int r2;
int r3 = (x != 0);
int r4 = (y != 0);
if (r1 != r3) {
   r2 = r4;
   z = 1;
} else {
   r2 = r1;
   z = (r2 == r4);
}
\end{verbatim}
\end{quote}
In effect, the \co{r2} load is merged with the \co{r1} load in
executions where \co{r1} is equal to \co{r3} or to \co{r4}, and it is
merged with the \co{r3} load in other executions.

@@

This paper focused solely on compilers, primarily those that do only
per-thread analysis and optimization.
It introduced the notion of quasi-volatile object accesses and gave
a definition for semantic dependency that is relative to a particular
C++ implementation.
It formulated a Fundamental Property of semantic dependency and proved
that implementations subject to some very mild restrictions on how
they treat atomic object accesses do have this property.
Finally, it demonstrated that satisfying the Fundamental Property
guarantees an implementation cannot give rise to OOTA executions.
